---
title: "Introduction to tangler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tangler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Welcome to tangler

`tangler` is an R package designed to create visually appealing tanglegrams using the `ggtree` and `ggplot2` ecosystem. This package primarily supports two and three phylogenies to visualize host-symbiont relationships, co-evolution, or simply comparing hierarchical clusterings.

### Getting Started

Load the necessary libraries:

```{r setup, eval=FALSE}
library(ggplot2)
library(ggtree)
library(dplyr)
library(tangler) 
```

### Main Functions

1.  **`common.tanglegram`**: Connects tips between two trees using a common categorical metadata column.
2.  **`simple.tanglegram`**: Connects tips between two trees based on a specific matching value within a metadata column.
3.  **`triple.tanglegram`**: Extends the capabilities of `common.tanglegram` to three phylogenies, allowing for the visualization of tripartite associations.

### Toy Example

Here's a quick example to generate random trees and plot a triple tanglegram:

```{r, eval=FALSE}
library(ape)

# Create 3 simple trees
set.seed(42)
t1 <- rtree(10)
t2 <- rtree(10)
t3 <- rtree(10)

t1$tip.label <- paste0("t", 1:10)
t2$tip.label <- paste0("t", 1:10)
t3$tip.label <- paste0("t", 1:10)

# Create metadata
meta <- data.frame(
  label = paste0("t", 1:10),
  category = sample(c("Group 1", "Group 2", "Group 3"), 10, replace = TRUE)
)

# Convert to ggtree objects and join metadata
tree1 <- ggtree(t1) %<+% meta
tree2 <- ggtree(t2) %<+% meta
tree3 <- ggtree(t3) %<+% meta

# 1. Simple Tanglegram (Filters for "Group 1")
simple.tanglegram(tree1, tree2, column = category, value = "Group 1")

# 2. Common Tanglegram
common.tanglegram(tree1, tree2, column = "category", t2_pad = 0.5)

# 3. Triple Tanglegram
triple.tanglegram(tree1, tree2, tree3, column = "category", t2_pad = 0.5, t3_pad = 0.5)
```
